---
title: "Assignment 2: Coding Basics"
author: "Ana Bishop"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r sequence}
#1.
sequence <- seq(1,100,4) # creating a sequence that starts at 1, ends at 100, and                                   increases by 4, and assigning it to the variable "sequence"
sequence # calling the sequence I created

#2. 
mean <- mean(sequence) #function that finds the mean of a variable
median <- median(sequence) #function that finds the median of a variable

mean
median

#3. 
mean>median 
# testing whether the mean is greater than the median. If it's not, 
#the output will be "FALSE" (which it was)
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

``` {r vectors}
students <- c("Anna", "Brad", "Charlie", "Susie")  #character vector
scores <- c(95, 60, 80, 40)  #numeric vector
pass <- c(TRUE, TRUE, TRUE, FALSE) #logical vector

class(students)  #shows what type of vector each vector is
class(scores)
class(pass)
```

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

```{r df}
test_information <- data.frame(students, scores, pass)
test_information
```

8.  Label the columns of your data frame with informative titles.

```{r colnames}
colnames(test_information) <- c("student_names","test_scores","pass_fail")
test_information
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:
Dataframes can hold multiple classes of data, while matrices can only hold one. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

```{r if/else}
## if/else
test_function <- function(x){
  if(x>=50) {
    TRUE
  }
  else{
    FALSE
  }
}

grades <- c(70, 40, 65, 20, 50, 90)
## test_function(grades)
# I had to make this^ a comment in order for the document to knit without errors
```

```{r ifelse}
## ifelse
passing_grade <- function(x){
  ifelse(x>=50, TRUE, FALSE)
}

passing_grade(grades)
```

11. Apply your function to the vector with test scores that you created in number 5.

```{r apply}
passing_grade(scores)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:
ifelse worked, while if/else did not. The if/else function returned an error stating that "the condition has length > 1". This means that the if/else function could not work with a vector that contained multiple grades, while the ifelse function could. The ifelse function knew to apply the conditional statement individually to each component of the vector.
